% ============================================================================
\section*{Appendix: Primitive Admission Proofs}
\label{sec:appendix-proofs}
\addcontentsline{toc}{section}{Appendix: Primitive Admission Proofs}

The Irreducibility Criterion of \Cref{sec:irreducibility} admits or rejects a candidate primitive by asking whether its cost can be expressed as a same-layer dataflow composition. The three cases below admit the MAC Unit and Attention and reject the Systolic Array.

\paragraph{Proof of irreducibility: MAC Unit (Ma).} At the Hardware layer, let $\text{Ma} = g(\text{Adder}, \text{Multiplier})$. A dataflow-graph composition yields:
\begin{align}
\label{eq:mac-proof}
&f\bigl(\text{Cost}_L(\text{Adder}),\, \text{Cost}_L(\text{Multiplier})\bigr) \notag \\
&\quad = N_{\text{cores}} \cdot f_{\text{clk}} \cdot 2 \;\approx\; 67\;\text{TFLOP/s}.
\end{align}
But $\text{Cost}_L(\text{Ma}) = \cvalint{\HhPeakTFLOPS}\;\text{TFLOP/s}$ (FP16 tensor cores). The $14.8\times$ gap arises from the tensor-core pipeline's $4{\times}4{\times}4$ warp-level matrix multiply, a microarchitectural optimization invisible at the arithmetic level. Since $\cvalint{\HhPeakTFLOPS} \neq 67$ under same-layer composition of adders and multipliers, the MAC is irreducible. \qed

\paragraph{Proof of irreducibility: Attention (At).} At the Algorithm layer, decompose attention into two Dense Dots and a row-wise softmax: $\text{At} = g(\text{Dd}_{QK}, \mathrm{softmax}, \text{Dd}_{PV})$. Each local operator has an independently expressible FLOP cost, but the composition creates the $n \times n$ score/probability residency induced by all-to-all token interaction. That peak live-memory term is a property of the attention pattern, not either dense dot in isolation:
\begin{equation}
\label{eq:attn-proof}
\text{Cost}_L(\text{At})_{\text{mem}} = \Theta(n^2).
\end{equation}
The $\Theta(n^2)$ residency term is absent from the local costs of $\text{Dd}_{QK}$, $\mathrm{softmax}$, and $\text{Dd}_{PV}$ considered separately. No same-layer dataflow graph over those local costs recovers the quadratic residency term without admitting Attention as the interaction primitive. Attention is irreducible. \qed

\paragraph{Proof of reducibility: Systolic Array.} A Systolic Array \citep{kung1982} is a regular grid of $N \times N$ MAC units connected by nearest-neighbor links. Its cost decomposes as:
\begin{align}
\label{eq:systolic-proof}
\text{Throughput}(\text{SA}_{N \times N}) &= N^2 \times \text{Throughput}(\text{Ma}) \\
\text{Latency}(\text{SA}_{N \times N}) &= (2N - 1) \times \text{Latency}(\text{Ma}) \notag \\
\text{Energy}(\text{SA}_{N \times N}) &= N^2 \times \text{Energy}(\text{Ma}) + E_{\text{link}} \notag
\end{align}
where $E_{\text{link}}$ is nearest-neighbor interconnect energy, itself a same-layer quantity. Assuming 100\% utilization and ignoring fill/drain effects, all three metrics are steady-state compositions of same-layer primitives. The Systolic Array is therefore reducible and rejected from the primitive vocabulary. \qed

\paragraph{Caveat on parameterization.} The result depends on the chosen primitive boundary. Treating interconnect energy, fill/drain effects, or utilization ceilings as emergent costs would rerun the proof at a finer granularity. The criterion's job is to make that choice explicit and replicable.
